---
layout: post
title: Tidy data and importing your data in R
details: Learn how to store your data in a tidy format to easily import it into R! Bring your own data!
location: FG423
date: 2016-06-16
time: '15:00'
topic: tidy data
packages: '"readr"'
video: "http://www.youtube.com/watch?v=qc75kuENzxQ"
---

In R, as in most statistical and programming languages, how your data is
structured and stored impacts how or whether you can do your analysis. The best
form for a dataset is in a tidy format. Tidy datasets are where each column
contains a variable measurements and each row contains a single observation at a
specific time. So, if you work with animals, each row would be the animal. For
cells, each row would be the cell plate/dish. For humans, each row would be an
individual. If you collect multiple time points on the animals/humans, they
would take up a new row for each measurement time.

This tidy format for the data makes it easier to analyze the dataset and to plot
it. It also makes it very easy to import into R. The best time to start using
this format for storing your data is right away! But, sometimes your data is
created by a machine or from the measurement instrument. In those cases, you'll
have to wrangle ('manipulate') the data into the appropriate tidy format before
analyzing it.

## Code used

Quick note: If you want to know more about a command (aka. function), use `?`. 
For instance, use `?install.packages` to learn more about the
`install.packages()` function.

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir = './')
```

```{r}
# install the package if you haven't already
# install.packages('readr')
# Load the readr package
library(readr)
```

We'll be using `read_csv()` from the `readr` package, which is similar to the
`read.csv()` function in base R but is slower and does more things that you
usually don't want. We can use `read_csv()` to also get a csv file, which is a
comma separated value file, from the internet. To get a better look at the csv
file, we can save it to a file by using `write_csv()`. The two datasets below
are fake datasets I made up.

```{r}
# This is an example of a messy dataset
web_link <- "http://codeasmanuscript.github.io/code-along/data/tidy-data-1.csv"
tidy_data <- read_csv(web_link)
# Save it into the data/ folder using write_csv
#write_csv(tidy_data, 'data/tidy-data-1.csv')

# Assign the tidy-data into the object ds using <-
# Import directly from a file:
ds <- read_csv('data/tidy-data-1.csv')

# Let's take a peek at this file too.
web_link2 <- "http://codeasmanuscript.github.io/code-along/data/tidy-data-2.csv"
read_csv(web_link2)
```

If you have a dataset that is tab separated, you can use `read_tsv()`. Since
this is a tab separated (or just plain text, txt, file), we need to save it to
the computer using `writeLines()`.

```{r}
read_tsv("http://codeasmanuscript.github.io/code-along/data/tidy-data-3.txt")
# Save it using:
# writeLines(
#     read_lines("http://codeasmanuscript.github.io/code-along/data/tidy-data-3.txt"), 
#     'data/tidy-data-3.txt')
```

These next two datasets are data Zhila in our lab collected for PROMISE, which is
the completely raw data file sent out from the machine she was using for measurements.
You can see when you open the `tidy-data-3.txt` file that there is a bunch of
stuff you probably don't need. So, tell R to not include these stuff when importing.
Each arguments (options) in the `read_tsv()` function is explained below:

- `comment` argument says that anything starting with `#` can be ignored
- `skip` says to skip the first line when importing
- `n_max` says to only import 8 lines *after* the header line (i.e. the column
names)
- `[-1:-2]` is called indexing (see `?'['`) and says to *exclude* (minus) the
first two columns
- `[c(-13, -14)]` (or `[-13:-14]`) takes the previous index results and says to
*exclude* columns 13 and 14

```{r}
ds <- read_tsv("http://codeasmanuscript.github.io/code-along/data/tidy-data-3.txt",
               comment = '#', skip = 1, n_max = 8)[-1:-2][c(-13, -14)]
# This adds a column called PlateRow to the ds object
ds$PlateRow <- 1:8
ds
```

In this case, to get a tidy dataset, we'll need to re-arrange the dataset using
another package.

```{r}
#install.packages('tidyr')
library(tidyr)
# gather takes a wide dataset can converts it to a long dataset. In this case
# I don't want PlateRow to be converted, just all other columns.
ds2 <- gather(ds, PlateColumn, Wavelength, -PlateRow)
ds2
```

And lastly, to look at this next dataset:

```{r}
# Save it using:
# writeLines(
#     read_lines("http://codeasmanuscript.github.io/code-along/data/tidy-data-4.csv"), 
#     'data/tidy-data-4.csv')

# [1:20, ] means to keep only the first 20 *rows*, in which a comma must follow
read_csv("http://codeasmanuscript.github.io/code-along/data/tidy-data-4.csv",
         skip = 22)[1:20, ]
```

# Resources

- [Tidy data paper](https://www.jstatsoft.org/index.php/jss/article/view/v059i10/v59i10.pdf)
- [A code heavier version of the Tidy Data paper](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
- [`readr` documentation](https://github.com/hadley/readr/blob/master/README.md)
- [Cheatsheat on aspects of working with data (including tidy data)](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
