---
title: Reproducible and dynamic manuscripts/theses in R
details: Learn how to use RStudio to write up your next paper or thesis.
location: FG423
date: 2016-11-30
time: '13:00'
topic: report generation
packages: c("rmarkdown", "knitr", "captioner")
permalink: /404
video: "https://www.youtube.com/watch?v=AfFRWRCAyHQ"
---

More detailed explanation of Rmarkdown can be found at the 
[RStudio Website](http://rmarkdown.rstudio.com/) or in the 
[pandoc website](http://pandoc.org/) (which is what R Markdown uses to convert
to other documents).

This was the code shown for the session:

## YAML Header

R Markdown files should have at the top a YAML header. YAML is just a series of
settings to use for the document, like title, author, what document to create
(Word, PDF, etc), where your bibliography is located etc.

```
---
title: "DNS session"
author: "Luke W. Johnston"
date: "November 30, 2016"
bibliography: sample.bib
csl: ama.csl
output: word_document
---
```

## Code chunk

Code chunks allow you to insert R output (like results, plots, tables, numbers,
etc) into the document. This is the primary strength of using R Markdown. Use
Ctrl-Alt-I to insert a code chunk when using RStudio.

## Headers

Inserting a header, such as Introductions or Methods, is easy:

```
# Header 1

## Header 2

### Header 3
```

## Bolding and italics

To **bold** use `**bold**`. To *italicize* use `*italicize*`.

## Some code used in the session:

```{r captions}
library(captioner)
tab_nums <- captioner(prefix = "Table")
tab1 <- tab_nums("tab1", "This is the caption for table 1. Looking at the regression from the swiss dataset.")
#tab1

fig_nums <- captioner(prefix = "Figure")
fig1 <- fig_nums("fig1", "This is a caption for figure 1, showing a scatter plot.")
```

If you want to see the results, see `r tab_nums("tab1", display = 'cite')`.

```{r table}
library(pander)
pander(lm(Fertility ~ Education, data = swiss), caption = tab1)
```

## Including Plots

You can also embed plots here (see `r fig_nums("fig1", display = "cite")`), for example:

```{r pressure, echo=FALSE, fig.cap=fig1}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

Inline numbers `r round(mean(swiss$Fertility), 2)`

# Citations

```{r, eval=FALSE}
library(curl)
curl_download("http://codeasmanuscript.github.io/code-along/misc/sample.bib", "sample.bib")
curl_download("https://raw.githubusercontent.com/citation-style-language/styles/master/american-medical-association.csl", "ama.csl")
```

According to certain authors blah blah `[@Lash2007a]`

These other authors say ... `[@An2013a; @Vandenbroucke2007a]`

It's good to have the last header be:

```
# References
```

... since the references will be inserted at the end.
