---
layout: post
title: Data visualization in R using ggplot2
details: Learn to plot your data into beautiful graphics!
location: FG423
date: 2016-06-30
time: '15:00'
topic: plotting
packages: '"ggplot2"'
last_modified: `r Sys.Date()`
notes: true
---

R has a beautiful set of plotting capabilities that allow it to produce
publication-quality graphs very easily and quickly. A commonly used package for
making graphs in R is called ggplot2. The reason it is so popular is because it
has great documentation, tutorials, and cheatsheets (see the Resources links on
the bottom), in addition to it being fairly easy to learn and use.

So let's get started. Load up the package:

```{r}
library(ggplot2)
```

Before starting, there is major assumption here for making plots: that your data
is already cleaned up and tidy, ready for plotting and analysis. If it isn't,
finish that part of your work first!

Let's look over two datasets. I'm using two here because one contains a lot of
continuous data (swiss) and the other contains more discrete data (mpg). Use
`colnames` to look at the column names of your dataset:

```{r}
colnames(swiss)
colnames(mpg)
```

Alright, let's do some simple plotting here. Standard plots include:

- Line graph
- Scatterplot
- Scatterplot with a regression/smoothing line
- Barplot
- Boxplot

The nice thing about ggplot2 is it is based on layers. You start with the base
`ggplot` function, and using `+` you add additional layers with the `geom_`
commands. Each type of layer ends in the type it is trying to create; so a line
graph would be `geom_line`, a scatterplot would be `geom_point`, a bar would be
`geom_bar`, and so on. Where you put the `geom_` in the layer will dictate where
it will be placed on the final plot. The other thing to use in ggplot2 is the
`aes` command, which stands for the aesthetics... or rather, what data and
values you *actually* want to plot. So `aes(x = Height, y = Weight)` would put
Height on the x-axis and Weight on the y-axis. Let's try it out.

## Common plots

### Line graph: Fertility by Agriculture

```{r lineplot}
ggplot(swiss, aes(x = Fertility, y = Agriculture)) +
    geom_line()
```

### Scatterplot: Education by Examination

```{r scatterplot}
ggplot(swiss, aes(x = Education, y = Examination)) +
    geom_point()
```

### Scatterplot with regression/smoothing lin: Education by Examination

Using [loess](https://en.wikipedia.org/wiki/Local_regression) smoothing line:

```{r smooth}
ggplot(swiss, aes(x = Education, y = Examination)) +
    geom_point() +
    geom_smooth()
```

... or a simple linear regression line:

```{r lmsmooth}
ggplot(swiss, aes(x = Education, y = Examination)) +
    geom_point() +
    geom_smooth(method = 'lm') # lm = linear model
```

### Barplot: Number of vehicle types (class)

```{r barplot}
ggplot(mpg, aes(x = class)) +
    geom_bar()
```

### Boxplot: Vehicle type (class) by highway miles/gallon (hwy)

```{r boxplot}
ggplot(mpg, aes(x = class, y = hwy)) +
    geom_boxplot()
```

### Sub-dividing up your plot: 

Let's plot drive type (4-wheel, front, rear) by highway mpg by number of
cylinders.

```{r facet_boxplot}
ggplot(mpg, aes(x = drv, y = hwy)) +
    geom_boxplot() +
    facet_grid(~ cyl)
```

There are dozens of types of layers (`geom_`) that you can use and the
documentation is incredible! So if there is a plot you want to make, you
definitely can do it in R!

## Customizing your plots:

Default, using density plot (which shows the distribution of a continuous
variable, useful for assessing skewness):

Note: `fill` tells ggplot2 how to fill in groups with a colour.

```{r densityplot}
ggplot(mpg, aes(x = hwy, fill = drv)) +
    geom_density(alpha = 0.3) # alpha = transparency
```

Adding a different colour (using the `scale_` group of commands; since `fill` is
used, it would be `scale_fill_` and since one of the colour palettes is called
`brewer`, it turns into `scale_fill_brewer`). You can see the different choices
for palettes by running `?scale_fill_brewer` to look at the help file.

```{r}
ggplot(mpg, aes(x = hwy, fill = drv)) +
    geom_density(alpha = 0.3) +
    scale_fill_brewer(palette = 'Pastel1')
```

And to customize individual features of the plot, you use `theme`. The `theme`
options are quite extensive, so if you want to look more into it, check out
`?theme` or the very detailed documentation
[here](http://docs.ggplot2.org/dev/vignettes/themes.html). There is a nice
graphic right above "Complete and incomplete theme objects" section, near the
bottom of the web document.

```{r}
ggplot(mpg, aes(x = hwy, fill = drv)) +
    geom_density(alpha = 0.3) +
    scale_fill_brewer(palette = 'Dark2') +
    theme(panel.border = element_blank(), # panel = background of the entire plot
          panel.background = element_blank(),
          panel.grid.major = element_line(colour = 'grey90'),
          legend.position = 'top',
          line = element_line(colour = 'black'),
          axis.line.y = element_line(colour = 'black'),
          axis.line.x = element_line(colour = 'black'))
```

# Resources:

- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) 
(in RStudio 'Help -> Cheatsheets -> Data Visualization')
- [ggplot2 documentation](http://docs.ggplot2.org/current/)
- [Tutorial from a course](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html)
- [DataCamp tutorials](https://www.datacamp.com/courses/data-visualization-with-ggplot2-1)
