---
layout: post
title: Basic statistics and R commands
details: Learn about common R and statistic commands.
location: FG423
date: 2016-10-14
time: '13:00'
topic: basic stats and R
packages: '"broom"'
video: "https://www.youtube.com/watch?v=eZbMtHAUMXQ"
---

This is the code that we went over during the session.

Outline of statistics code we used:

- mean: mean
- sd: standard deviation
- median: median
- IQR: Interquartile range
- cor: correlation
- cor.test: correlation with p-value
- aov: ANOVA
- glm: linear regression

## Loading the data!

Import the data from the Code As Manuscript website to use in this session.

```{r}
ds <- read.csv("http://codeasmanuscript.org/states_data.csv")
names(ds)
```

## Simple descriptive statistics

The `$` tells R to take a column from the dataset. So `ds$Population` tells R to
take the Population column from the ds object.

```{r}
mean(ds$Population)
sd(ds$Population)

round(median(ds$LifeExp), 1)
IQR(ds$LifeExp)
```

You can use R code inline too:

    The median of Life Expectancy in the States dataset is: `r round(median(ds$LifeExp),1)`

The median of Life Expectancy in the States dataset is: `r round(median(ds$LifeExp),1)`

You can put the results of the statistical output into an object called `Corr` (R is *case-sensitive*). And if we load the `broom` package, we can use the `tidy()` function 

```{r}
cor(ds$Population, ds$LifeExp)
Corr <- cor.test(ds$Population, ds$LifeExp)
library(broom)
Corr2 <- tidy(Corr)
```

Again, you can use inline R code to convert to text:

    The correlation is `r Corr2$estimate` and it is not significant (p=`r Corr2$p.value`).

The correlation is `r Corr2$estimate` and it is not significant (p=`r Corr2$p.value`).

Running an ANOVA uses a slightly different syntax: the `~` (tilde) is used in R
to indicate that it is a formula. So in this case, you are seeing the role of
`Division` (the X variable as a factor/discrete) on `Population` (the y variable
as a continuous variable).

```{r}
AOV <- aov(Population ~ Division, data = ds)
summary(AOV)
AOV2 <- tidy(AOV)
```

    There is a significant difference in Population between Divisions (p=`r AOV2$p.value[1]`). The table is below:

There is a significant difference in Population between Divisions (p=`r AOV2$p.value[1]`). The table is below:

You can put the ANOVA results into a table too!

```{r}
knitr::kable(AOV2)
```

Linear regression uses the same notation as ANOVA with the `~`. In this case, since this is a linear regression you need to use a Gaussian distribution, as `LifeExp` (the y variable) is continuous.

```{r}
fit <- glm(LifeExp ~ Income, data = ds, family = gaussian)
summary(fit)
fit2 <- tidy(fit)
knitr::kable(fit2)
```

You can use math formulas inside R Markdown files too:

    $$y = Xb + e$$
    
The regression formula is similar to the formula style in R with the `~`.
    
$$y = Xb + e$$
    
If you want to go through each step of how linear regression works, I wrote a 
[blog on understanding linear regression](http://blog.lukewjohnston.com/variance-linear-regression). Check it.

We can add covariates (confounders, etc) to the model, just like the math formula.

$$y = x1b + x2b + e$$

And in R:

```{r}
fit_covar <- glm(LifeExp ~ Income + Population, data = ds, family = gaussian)
summary(fit_covar)
fit_covar2 <- tidy(fit_covar)
knitr::kable(fit_covar2)
```

If you want to select just certain variables to display in a table, you can more easily do that using the package `dplyr`:

```{r}
library(dplyr)
fit_covar2 %>% 
    select(Variable = term, 
           Beta = estimate, 
           P = p.value) %>% 
    knitr::kable()
```

