---
layout: post
title: Organizing and simplifying code using functions
details: Come learn how to make your code reusable and easier to use with functions.
location: FG423
date: 2017-02-01
time: '13:00'
topic: reusing your code
output: slidy_presentation
video: "https://www.youtube.com/watch?v=a0UJHS7Kr3o"
---

# How to reuse your code... Use functions!

- Use a piece of code more than once? Create a function!
- Have a bit of code that is complex or fairly complicated? Create functions!

## Why use functions:

- Makes your code more readable
- Let's you reuse complex or repetitive code easily
- Uses the DRY principle: Don't Repeat Yourself

# What is a function:

- An encapsulated piece of code
- Takes an input or set of inputs, and outputs a single object
- R is a functinal programming language: Every action is a function in R
    - ie. `mean()`, `plot()`, `[`, `<-`, etc.

# Creating a function

- Each function is composed of at least three to four parts:
    - Assignment `<-` as an object (optional)
    - The `function` call and if needed the arguments (aka options/settings)
    - The code within the function
    - Final object/data to output (the `return`)
- Make your code more readable, name your function descriptively:
    - `scatter_plot` **not** e.g. `someplot`
    - `extract_pvalue` **not** e.g. `p` or `value`
- Be descriptive with the code in the function too

# A function looks like:

```{r, eval=FALSE}
# Anonymous function
function(arg1, arg2) {
    #...code content and action...
    return(output)
}

# Named function
function_name <- function(arg1, arg2) {
    #...code content and action...
    return(output)
}
```

# A very simple example:

```{r}
Add <- function(value1, value2) {
    output <- value1 + value2
    return(output)
}

Add(1, 2) # should be 
```

# Document your functions

Use `Ctrl-Shift-Alt-R` with cursor in function when using RStudio (only used in
`.R` files, not `.Rmd` files)

Should look like:

```{r, eval=FALSE}
#' Informative title
#' 
#' More detailed description.
#'
#' @param value1 explanation
#' @param value2 explanation
#'
#' @return What it outputs
#'
#' @examples
#' Add(1,2)
```

# if else in functions

- Control how your function works using `if else`:

```{r, eval=FALSE}
if (condition) {
    true condition
} else {
    false condition
}

if (condition) {
    true condition
} else if (another_condition) {
    true condition
} else {
    false condition
}
```

# Using functions from other packages in your function

Sometimes you'll come across this `::`. For instance:

```{r, eval=FALSE}
dplyr::select(iris, Sepal.Length)
```

- This tells R to take the function `select` from the package `dplyr`. 
- Better to use this style *within* a function rather than use `library()`
    - Using `library()` within a function is bad practice
    - Let's you know which commands come from which package
    - Better documents your function
    - Makes it more readable and transparent
    
# Code used in session:

```{r}
# Build up function without `::` to check that it works
library(ggplot2)
gg_scatterplot <- function(data, x, y) {
    scatter_plot <- ggplot(data, aes_string(x, y)) +
        geom_point() +
        theme_bw() +
        theme(panel.border = element_blank(),
              panel.grid = element_blank(),
              axis.line.x = element_line(colour = 'black'),
              axis.line.y = element_line(colour = 'black')
              )
    return(scatter_plot)
}

gg_scatterplot(iris, 'Sepal.Length', 'Sepal.Width')

# It works! Now switch to using `::`
gg_scatterplot <- function(data, x, y) {
    scatter_plot <- ggplot2::ggplot(data, ggplot2::aes_string(x, y)) +
        ggplot2::geom_point() +
        ggplot2::theme_bw() +
        ggplot2::theme(panel.border = ggplot2::element_blank(),
              panel.grid = ggplot2::element_blank(),
              axis.line.x = ggplot2::element_line(colour = 'black'),
              axis.line.y = ggplot2::element_line(colour = 'black')
              )
    return(scatter_plot)
}

gg_scatterplot(iris, 'Sepal.Length', 'Sepal.Width')
gg_scatterplot(iris, 'Sepal.Length', 'Petal.Length')
```

```{r}
# Using if else conditions
anova_results <- function(data, factor_var, cont_var, tidied = TRUE) {
    Y <- factor(data[[factor_var]])
    x <- as.numeric(data[[cont_var]])
    if (tidied) {
        model <- broom::tidy(aov(x ~ Y))
    } else {
        model <- aov(x ~ Y)
    }
    return(model)
}

anova_results(iris, 'Species', 'Sepal.Length')
anova_results(iris, 'Species', 'Petal.Length')
anova_results(iris, 'Species', 'Petal.Length', tidied = FALSE)

# Tidied version allows you to easily access values for the anova results.
fit <- anova_results(iris, 'Species', 'Petal.Length',
              tidied = TRUE)
fit$p.value
fit$sumsq

# Compared to...:
fit1 <- anova_results(iris, 'Species', 'Petal.Length',
              tidied = FALSE)
fit1$p.value
```

```{r}
# Simple example of using anonymous functions:
processing_data <- function(data, char_var, fun) {
    data$NewVar <- fun(data[[char_var]])
    return(data)
}

# Use the anonymous function as an argument for the `fun` parameter of the processing_data function.
processing_data(head(iris, 20), 'Species', function(x) {
    x <- gsub('setosa', 'Set', x)
    x <- as.factor(x)
    return(x)
})
```
